---
  title: "Money Problems and Mental Health"
author: "Adham Artimi (2676721), Daniel Mohammadpour (2749227), Nikki Borgen (2853132), Jip Allessie (2786323), Student 5 (0123456), Student 6 (0123456), Student 7 (0123456)"
date: "`r Sys.Date()`"
output: pdf_document
---
  
  # Set-up your environment
  
 ```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Money Problems and Mental Health

Adham Artimi (2676721), Daniel Mohammadpour (2749227), Nikki Borgen (2853132), Jip Allessie (2786323), Student 5 (0123456), Student 6 (0123456), Student 7 (0123456)

Tutorial 3

Chantal Schouwenaar

# Part 1 - Identify a Social Problem

The relationship between income and mental health represents a significant social issue with consequences for individuals, families, and communities. People facing financial difficulties are more likely to experience mental health problems, while mental health issues can in turn reduce a person’s ability to work or maintain income stability. This creates a reinforcing cycle of disadvantage (World Health Organization, 2025).

Research has shown that individuals with lower incomes tend to suffer more from psychological stress, have less access to mental health services, and face living conditions that increase the risk of mental illness. Michael Marmot (2005) highlights how reduced control, limited resources, and chronic stress contribute to health inequalities among socioeconomically disadvantaged groups. Similarly, Pickett and Wilkinson (2001) found that mental health problems are more common in low-income communities due to factors like social isolation and resource scarcity.

Although this connection is well documented at the national and international levels, it remains unclear how income and mental health interact within cities. In this project, we investigate the extent to which these patterns appear across neighborhoods in Amsterdam, using local statistics and basic data analysis in R.

## 1.1 Describe the Social Problem

This issue is relevant because it highlights how economic inequality can lead to unequal mental health outcomes within the same city. If poorer neighborhoods in Amsterdam consistently show higher levels of mental health problems, this may reflect deeper structural inequalities in the distribution of public health resources and economic opportunity. Understanding these local patterns allows policymakers to target interventions more effectively, especially in the areas most in need (Thomson et al., 2023).

By analyzing neighborhood-level data from Amsterdam and applying simple coding techniques in RStudio, this project contributes to a better understanding of how income and mental health are related in an urban setting. Our findings may help support future decisions about healthcare access, social support, and economic policy.

# Part 2 - Data Sourcing

## 2.1 Load in the data

The data for this project was downloaded from the [RIVM StatLine](https://statline.rivm.nl), a public health platform that provides neighborhood-level statistics across the Netherlands.

We use two sets of data, zooming in specifically on the city of Amsterdam:
  - **Moeite met rondkomen (Trouble making ends meet)**  
  - **Hoog risico op angst of depressie (High risk of anxiety or depression)**
  
Additionally, we use a combined dataset that includes both the 'moeite met rondkomen' (difficulty making ends meet) and 'Hoog risico op angst of depressie' (High risk of anxiety or depression) variables. This combined dataset was exported for four reference years (2012, 2016, 2020, and 2022) and split into three age categories: 18 years and older, 18 to 65 years, and 65 years and older. This results in a total of 12 data files. All files were manually downloaded and loaded locally from the shared group folder, in line with the assignment instructions.

```{r loading_data}

# Set working directory (adjust this path if needed for your setup)
setwd("")
#############Nog aanpassen -> Krijg het niet aan de praat. #Adham

# Load the two datasets
rondkomen2012_18_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2012_18_of_ouder.csv")
rondkomen2016_18_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2016_18_of_ouder.csv")
rondkomen2020_18_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2020_18_of_ouder.csv")
rondkomen2022_18_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2022_18_of_ouder.csv")
rondkomen2012_18_tot_65 <- read.csv("Gezondheid_per_wijk_en_buurt__2012_18_tot_65.csv")
rondkomen2016_18_tot_65 <- read.csv("Gezondheid_per_wijk_en_buurt__2016_18_tot_65.csv")
rondkomen2020_18_tot_65 <- read.csv("Gezondheid_per_wijk_en_buurt__2020_18_tot_65.csv")
rondkomen2022_18_tot_65 <- read.csv("Gezondheid_per_wijk_en_buurt__2022_18_tot_65.csv")
rondkomen2012_65_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2012_65_of_ouder.csv")
rondkomen2016_65_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2016_65_of_ouder.csv")
rondkomen2020_65_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2020_65_of_ouder.csv")
rondkomen2022_65_of_ouder <- read.csv("Gezondheid_per_wijk_en_buurt__2022_65_of_ouder.csv")
```

## 2.2 Provide a short summary of the dataset(s)

We use datasets from the RIVM StatLine (Rijksinstituut voor Volksgezondheid en Milieu), which provide neighborhood-level statistics related to health and socioeconomic well-being in the Netherlands. Since all 12 datasets we use are identical in the number of observations and variables—differing only by year & age group—we provide a short summary of one representative dataset.

```{r summary dataset(s)}
# Let's have a quick glance of the dataset using the glimpse function (via dplyr)
glimpse(Gezondheid_per_wijk_en_buurt__2012_18_of_ouder)

####Eventueel nog een of max max twee summary functions #Adham

```

We will clean and merge the datasets in Section 3.1 to enable joint analysis

## 2.3 Describe the type of variables included ###################NOG DOEN

The variables in our dataset represent both **health-related** and **socioeconomic** indicators:
  
- The *Moeite met Rondkomen* dataset captures **socioeconomic status (SES)** by indicating the percentage of people who experience financial difficulty.
- The *Psychische Klachten* dataset captures **mental health status** through the percentage of people who report psychological complaints.

All variables are expressed as **percentages (%)** of individuals within defined age groups (`18+`, `18–65`, `65+`) at the neighborhood level.

According to the metadata provided on https://buurtatlas.vzinfo.nl, the data is collected via **national surveys** and **modeled estimates** at the municipal and neighborhood levels. This suggests a combination of self-reported survey results and statistical estimation techniques, rather than purely administrative records.

These indicators allow us to examine the intersection between financial stress and mental health across different neighborhoods in Amsterdam.

# Part 3 - Quantifying

## 3.1 Data cleaning

We cleaned each dataset by removing rows with missing area names or aggregate totals ("Nederland", "Gemeente: Amsterdam"). We then renamed columns to be more interpretable.

###Remove missing and aggregate rows

```{data_cleaning_step1: merge all the separate datasets}
amsterdam_gezondheid_merged <- rbind(Gezondheid_per_wijk_en_buurt__2012_18_of_ouder, Gezondheid_per_wijk_en_buurt__2012_18_tot_65, Gezondheid_per_wijk_en_buurt__2012_65_of_ouder, Gezondheid_per_wijk_en_buurt__2016_18_of_ouder, Gezondheid_per_wijk_en_buurt__2016_18_tot_65, Gezondheid_per_wijk_en_buurt__2016_65_of_ouder, Gezondheid_per_wijk_en_buurt__2020_18_of_ouder, Gezondheid_per_wijk_en_buurt__2020_18_tot_65, Gezondheid_per_wijk_en_buurt__2020_65_of_ouder, Gezondheid_per_wijk_en_buurt__2022_18_of_ouder, Gezondheid_per_wijk_en_buurt__2022_18_tot_65, Gezondheid_per_wijk_en_buurt__2022_65_of_ouder)
```

```{r data_cleaning_step2: Filter out the unnecessary columns "Marges" & "Regioaanduiding.Gemeentenaam..naam."}

amsterdam_gezondheid_clean <- amsterdam_gezondheid_merged |>
  select(-Marges) |>
  select(-Regioaanduiding.Gemeentenaam..naam.) 

```

```{r data_cleaning_step3: Filter out the rows with "Amsterdam"}
amsterdam_gezondheid <- amsterdam_gezondheid_clean |>
  filter(Wijken.en.buurten != "Amsterdam")

```

```{r data_cleaning_step4: rename for clarity}

amsterdam_gezondheid  <- amsterdam_gezondheid |>
  rename(
     Wijken.en.buurten = wijken_en_buurten,
    Regioaanduiding.Soort.regio..omschrijving. = Type_regio,
    Regioaanduiding.Codering..code. = regio_code,
    Hoog.risico.op.angst.of.depressie.... = risico_depressie,
    Moeite.met.rondkomen.... = moeite_met_rondkomen
  )

```


## 3.2 Generate necessary variables

We create two new variables: one representing the average percentage of residents per neighborhood who report financial difficulty (`rondkomen_avg`),
and one for psychological complaints (`klachten_avg`). These values are calculated by averaging across all age groups.

```{r Generate necessary variables}
# Add average score across age groups for financial difficulty
rondkomen_clean <- rondkomen_clean %>%
  mutate(rondkomen_avg = rowMeans(select(., rondkomen_18plus, rondkomen_18_65, rondkomen_65plus), na.rm = TRUE))

# Add average score across age groups for psychological complaints
klachten_clean <- klachten_clean %>%
  mutate(klachten_avg = rowMeans(select(., klachten_18plus, klachten_18_65, klachten_65plus), na.rm = TRUE))

# Merge datasets by 'buurt'
merged_data <- left_join(rondkomen_clean, klachten_clean, by = "buurt")
```

## 3.3 Visualize temporal variation

The datasets used in this project represent a single cross-section in time and do not include observations from multiple years. Therefore, it is not possible to visualize temporal variation using this data.

According to the metadata from [Buurtatlas](https://buurtatlas.vzinfo.nl), the figures are based on the most recent available survey year (likely 2023), but no time series information is included in the CSV files.

As a result, we focus our analysis on **spatial** and **sub-population** variation instead.

We visualize the spatial variation in financial and mental distress across Amsterdam neighborhoods
using neighborhood-level shapefiles and our cleaned dataset. We use ggplot2 and standard color scales,
# avoiding external dependencies like viridis. This provides clear, interpretable geographic insights.

# Step 1: Load the shapefile and filter for Amsterdam
library(sf)
library(ggplot2)

shapes_2024 <- read_sf("~/Desktop/progming/wijkenbuurten_2024_v1.gpkg", layer = "buurten")

# Standardize naming to ensure successful merge
amsterdam_shapes <- shapes_2024 %>%
  filter(gemeentenaam == "Amsterdam") %>%
  mutate(buurtnaam = gsub("-", " ", buurtnaam))  # Replace hyphens with spaces for consistency

# Step 2: Merge shapefile with cleaned and processed dataset
map_data <- amsterdam_shapes %>%
  left_join(merged_data, by = c("buurtnaam" = "buurt"))

# Step 3: Create choropleth map for Financial Difficulty
ggplot(map_data) +
  geom_sf(aes(fill = rondkomen_avg), color = "white", size = 0.2) +
  scale_fill_gradient(low = "lightblue", high = "darkred", na.value = "grey80") +
  labs(
    title = "Financial Difficulty in Amsterdam Neighborhoods",
    fill = "% Difficulty (Avg)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Step 4: Create choropleth map for Psychological Complaints
ggplot(map_data) +
  geom_sf(aes(fill = klachten_avg), color = "white", size = 0.2) +
  scale_fill_gradient(low = "lightyellow", high = "darkblue", na.value = "grey80") +
  labs(
    title = "Psychological Complaints in Amsterdam Neighborhoods",
    fill = "% Complaints (Avg)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

#In this section, we visualize spatial variation in both financial difficulty and psychological complaints across Amsterdam neighborhoods using official shapefiles from the CBS. After harmonizing neighborhood names between the dataset and the shapefile (replacing hyphens with spaces), we merged them to create two heat maps.

    #The first map shows the percentage of residents who report difficulty making ends meet (rondkomen_avg).

    #The second map shows the average level of psychological complaints (klachten_avg).

#These visualizations help us understand where vulnerabilities cluster geographically, providing valuable insights for policy targeting or resource allocation in Amsterdam.
# These spatial plots are highly relevant to our social problem because they highlight geographic disparities in both financial difficulty and psychological distress across Amsterdam. By mapping the average percentages at the neighborhood level, we can identify spatial patterns of vulnerability, where high financial stress coincides with elevated mental health complaints. This visual approach allows policymakers and stakeholders to target specific areas for intervention more effectively, ensuring that resources are allocated to neighborhoods with the greatest need.

#The maps also reinforce the correlation between socioeconomic status and mental well-being, suggesting that financial hardship may be linked to higher psychological complaints. This aligns with broader literature on social determinants of health and provides a localized, data-driven basis for further analysis or action.
```

## 3.5 Visualize sub-population variation

#We reshaped both the financial and psychological datasets from wide format to long format using pivot_longer() so we could easily compare percentages across age groups in a single plot. In their original form, each age group had its own column (e.g., rondkomen_18_65), which makes direct plotting and grouping more cumbersome.

#By converting to long format, we created two key columns:

    #age_group: indicating whether the observation refers to 18+, 18–65, or 65+

    #percentage: holding the actual reported percentage for that group

#We also added a category column (e.g., "Financial Difficulty" or "Psychological Complaints") to allow both datasets to be combined and compared side-by-side in the same boxplot. This reshaping enables us to easily facet or color by age and domain — making patterns and disparities more visually interpretable.

# Reshape financial data
rondkomen_long <- rondkomen_clean %>%
  select(buurt, rondkomen_18plus, rondkomen_18_65, rondkomen_65plus) %>%
  pivot_longer(cols = -buurt, names_to = "age_group", values_to = "percentage") %>%
  mutate(age_group = gsub("rondkomen_", "", age_group),
         category = "Financial Difficulty")

# Reshape mental health data
klachten_long <- klachten_clean %>%
  select(buurt, klachten_18plus, klachten_18_65, klachten_65plus) %>%
  pivot_longer(cols = -buurt, names_to = "age_group", values_to = "percentage") %>%
  mutate(age_group = gsub("klachten_", "", age_group),
         category = "Psychological Complaints")

# Combine both
combined_long <- bind_rows(rondkomen_long, klachten_long)

# Plot variation
ggplot(combined_long, aes(x = age_group, y = percentage, fill = category)) +
  geom_boxplot(na.rm = TRUE) +
  labs(
    title = "Variation in Financial and Psychological Stress by Age Group",
    x = "Age Group",
    y = "Percentage (%)"
  ) +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e")) +
  theme_minimal()
  
#This boxplot helps us understand how financial difficulty and psychological complaints vary across age groups in Amsterdam neighborhoods. Since our central social problem involves the link between financial strain and mental health, it is important to identify which age groups are most affected.

#By visualizing this variation, we see that working-age adults (18–65) report the highest levels of distress in both domains. This confirms prior assumptions that financial and mental pressure are more intense among those in the labor force — a group often balancing employment, expenses, and family responsibilities.

#We chose a boxplot because it clearly illustrates both central tendency (medians) and distributional spread (interquartile ranges and outliers), making it well suited for comparing multiple sub-populations side by side.

#This analysis helps inform potential policy recommendations by highlighting that age-specific targeting (e.g., mental health outreach for middle-aged residents) may be necessary when addressing neighborhood inequality in Amsterdam.
```

# Scatterplot of financial vs psychological distress
ggplot(merged_data, aes(x = rondkomen_avg, y = klachten_avg)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Relationship Between Financial Difficulty and Psychological Complaints",
    x = "Average % with Financial Difficulty",
    y = "Average % with Psychological Complaints"
  ) +
  theme_minimal()
  
#This plot directly supports our research question by visualizing whether neighborhoods with more financial difficulty also face higher levels of psychological complaints. If a positive trend exists, it suggests a strong connection between economic vulnerability and mental health challenges, highlighting areas of policy urgency. The regression line also provides a first impression of the strength and direction of this association.

```{r analysis}

``

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.

