---
title: "Money Problems and Mental Health"
author: "Adham Artimi (2676721), Daniel Mohammadpour (2749227), Nikki Borgen (2853132), Jip Allessie (2786323), Florian Hololtcheff
 (2790532), Raouf Dönmez (2870124), Adam Affara (2800635)"
date: "`r Sys.Date()`"
output: pdf_document
---
  # Set-up your environment
  
```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("lubridate")
install.packages("sf")
install.packages("tinytex")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(lubridate)
require(sf)
require(tmap)
library(tinytex)
```

# Money Problems and Mental Health

Adham Artimi (2676721), Daniel Mohammadpour (2749227), Nikki Borgen (2853132), Jip Allessie (2786323), Student 5 (0123456), Student 6 (0123456), Student 7 (0123456)

Tutorial 3; Chantal Schouwenaar

# Part 1 - Identify a Social Problem

The relationship between income and mental health represents a significant social issue with consequences for individuals, families, and communities. People facing financial difficulties are more likely to experience mental health problems, while mental health issues can in turn reduce a person’s ability to work or maintain income stability. This creates a reinforcing cycle of disadvantage (World Health Organization, 2025).

Research has shown that individuals with lower incomes tend to suffer more from psychological stress, have less access to mental health services, and face living conditions that increase the risk of mental illness. Michael Marmot (2005) highlights how reduced control, limited resources, and chronic stress contribute to health inequalities among socioeconomically disadvantaged groups. Similarly, Pickett and Wilkinson (2001) found that mental health problems are more common in low-income communities due to factors like social isolation and resource scarcity.

Although this connection is well documented at the national and international levels, it remains unclear how income and mental health interact within cities. In this project, we investigate the extent to which these patterns appear across neighborhoods in Amsterdam, using local statistics and basic data analysis in R.

## 1.1 Describe the Social Problem

This issue is relevant because it highlights how economic inequality can lead to unequal mental health outcomes within the same city. If poorer neighborhoods in Amsterdam consistently show higher levels of mental health problems, this may reflect deeper structural inequalities in the distribution of public health resources and economic opportunity. Understanding these local patterns allows policymakers to target interventions more effectively, especially in the areas most in need (Thomson et al., 2023).

By analyzing neighborhood-level data from Amsterdam and applying simple coding techniques in RStudio, this project contributes to a better understanding of how income and mental health are related in an urban setting. Our findings may help support future decisions about healthcare access, social support, and economic policy.

# Part 2 - Data Sourcing

## 2.1 Load in the data

The data for this project was downloaded from the [RIVM StatLine](https://statline.rivm.nl), a public health platform that provides neighborhood-level statistics across the Netherlands.

We use two sets of data, zooming in specifically on the city of Amsterdam:

  - **Moeite met rondkomen (Trouble making ends meet)**  
  - **Hoog risico op angst of depressie (High risk of anxiety or depression)**
  
Additionally, we use a combined dataset that includes both the 'moeite met rondkomen' (difficulty making ends meet) and 'Hoog risico op angst of depressie' (High risk of anxiety or depression) variables. This combined dataset was exported for four reference years (2012, 2016, 2020, and 2022) and split into three age categories: 18 years and older, 18 to 65 years, and 65 years and older. This results in a total of 12 data files. All files were downloaded from the shared group folder, in line with the assignment instructions.

```{r loading_data}

# Load the datasets
Gezondheid_per_wijk_en_buurt__2012_18_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2012_18_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2016_18_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2016_18_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2020_18_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2020_18_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2022_18_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2022_18_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2012_18_tot_65 <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2012_18_tot_65.csv")
Gezondheid_per_wijk_en_buurt__2016_18_tot_65 <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2016_18_tot_65.csv")
Gezondheid_per_wijk_en_buurt__2020_18_tot_65 <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2020_18_tot_65.csv")
Gezondheid_per_wijk_en_buurt__2022_18_tot_65 <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2022_18_tot_65.csv")
Gezondheid_per_wijk_en_buurt__2012_65_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2012_65_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2016_65_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2016_65_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2020_65_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2020_65_of_ouder.csv")
Gezondheid_per_wijk_en_buurt__2022_65_of_ouder <- read.csv2("https://raw.githubusercontent.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4/refs/heads/main/data/Gezondheid_per_wijk_en_buurt__2022_65_of_ouder.csv")
```

## 2.2 Provide a short summary of the dataset(s)

We use datasets from the RIVM StatLine (Rijksinstituut voor Volksgezondheid en Milieu), which provide neighborhood-level statistics related to health and socioeconomic well-being in the Netherlands. Since all 12 datasets we use are identical in the number of observations and variables—differing only by year & age group—we provide a short summary of one representative dataset.

```{r summary dataset(s)}
# Let's have a quick glance of the dataset using the glimpse (via dplyr) & head function
glimpse(Gezondheid_per_wijk_en_buurt__2012_18_of_ouder)
head(Gezondheid_per_wijk_en_buurt__2012_18_of_ouder)

```

This dataset contains 579 rows and 9 variables. Each row represents a specific neighborhood or district, and the variables include the year of data collection (Perioden), age group (Leeftijd), and values like the percentage of people at high risk for anxiety or depression and the percentage of people reporting difficulty making ends meet. These indicators are relevant for our research question, as they help us explore whether there is a possible relationship between financial difficulty and mental health at the neighborhood level.

## 2.3 Describe the type of variables included The variables in our 
dataset represent both health-related and socioeconomic indicators:

- The **Moeite met Rondkomen** variable reflects socioeconomic stress, measured by the percentage of individuals who report difficulty making ends meet.

- The **Hoog Risico op Angst of Depressie** variable reflects mental health vulnerability, measured by the percentage of individuals estimated to be at high risk for anxiety or depression.

These two variables are the focus of our analysis, as they relate directly to our research question about the connection between financial strain and mental health.

All variables are expressed as percentages (%) of individuals within defined age groups (18+, 18–65, or 65+) at the neighborhood (wijk/buurt) level.

In addition to the two main variables, the dataset includes structural variables such as:

- Wijken.en.buurten: the name of the neighborhood or district,
- Perioden: the year of data collection,
- Leeftijd: the age group,
- RegioS: the statistical code for the neighborhood or district.

According to the metadata available at RIVM Statline, the health-related indicators are based on national surveys combined with model-based estimates.

These variables provide the structure and context necessary to filter and analyze the data by location, year, and age group. This helps us focus on relevant subpopulations and make meaningful comparisons across Amsterdam’s neighborhoods.

# Part 3 - Quantifying

## 3.1 Data cleaning

We cleaned each dataset by removing rows with missing area names or aggregate totals ("Nederland", "Gemeente: Amsterdam"). We then renamed columns to be more interpretable.

###Remove missing and aggregate rows

```{r data_cleaning_step1: merge all the separate datasets}

amsterdam_gezondheid_merged <- rbind(Gezondheid_per_wijk_en_buurt__2012_18_of_ouder, Gezondheid_per_wijk_en_buurt__2012_18_tot_65, Gezondheid_per_wijk_en_buurt__2012_65_of_ouder, Gezondheid_per_wijk_en_buurt__2016_18_of_ouder, Gezondheid_per_wijk_en_buurt__2016_18_tot_65, Gezondheid_per_wijk_en_buurt__2016_65_of_ouder, Gezondheid_per_wijk_en_buurt__2020_18_of_ouder, Gezondheid_per_wijk_en_buurt__2020_18_tot_65, Gezondheid_per_wijk_en_buurt__2020_65_of_ouder, Gezondheid_per_wijk_en_buurt__2022_18_of_ouder, Gezondheid_per_wijk_en_buurt__2022_18_tot_65, Gezondheid_per_wijk_en_buurt__2022_65_of_ouder)
```

```{r data_cleaning_step2: Filter out the unnecessary columns "Marges" & "Regioaanduiding.Gemeentenaam..naam."}

amsterdam_gezondheid_clean <- amsterdam_gezondheid_merged |>
  select(-Marges) |>
  select(-Regioaanduiding.Gemeentenaam..naam.) 

```

```{r data_cleaning_step3: Filter out the rows with "Amsterdam"}
amsterdam_gezondheid <- amsterdam_gezondheid_clean |>
  filter(Wijken.en.buurten != "Amsterdam")

```

```{r data_cleaning_step4: rename for clarity}

amsterdam_gezondheid  <- amsterdam_gezondheid |>
  rename(
     wijken_en_buurten = Wijken.en.buurten,
    Type_regio = Regioaanduiding.Soort.regio..omschrijving.,
    regio_code = Regioaanduiding.Codering..code.,
    risico_depressie = Hoog.risico.op.angst.of.depressie....,
    moeite_met_rondkomen = Moeite.met.rondkomen....
  )

```

```{r data_cleaning_step5: set variables as numeric} 

amsterdam_gezondheid$risico_depressie = as.numeric(amsterdam_gezondheid$risico_depressie, na.rm = TRUE)

amsterdam_gezondheid$moeite_met_rondkomen = as.numeric(amsterdam_gezondheid$moeite_met_rondkomen, na.rm = TRUE)

```

```{r data_cleaning_step6: set periode as year} 

amsterdam_gezondheid$Perioden = ymd(amsterdam_gezondheid$Perioden, truncated = 2L)

```


## 3.2 Generate necessary variables

Variable 1

```{r gen_var1: Averages}

amsterdam_gezondheid <- amsterdam_gezondheid |>
  group_by(Perioden, Leeftijd) |>
  mutate(
    avg_risico_depressie = mean(risico_depressie, na.rm = TRUE),
    avg_moeite_rondkomen = mean(moeite_met_rondkomen, na.rm = TRUE)
  )

```

Variable 2

```{r gen_var2}

# First, calculate the percentiles (1–5) for "moeite met rondkomen" (financial stress). We assign each neighborhood to a quintile (1 = lowest stress, 5 = highest stress) and convert the numeric quintile into a descriptive label
amsterdam_gezondheid <- amsterdam_gezondheid |>
  mutate(
    rondkomen_percentile = ntile(moeite_met_rondkomen, 5),
    rondkomen_group = case_when(
      rondkomen_percentile == 1 ~ "Very Low",
      rondkomen_percentile == 2 ~ "Low",
      rondkomen_percentile == 3 ~ "Moderate",
      rondkomen_percentile == 4 ~ "High",
      rondkomen_percentile == 5 ~ "Very High",
      TRUE ~ NA_character_
    )
  )

# We reepeat the same logic for "risico op depressie" (mental health complaints)
amsterdam_gezondheid <- amsterdam_gezondheid %>%
  mutate(
    depressie_percentile = ntile(risico_depressie, 5),
    depressie_group = case_when(
      depressie_percentile == 1 ~ "Very Low",
      depressie_percentile == 2 ~ "Low",
      depressie_percentile == 3 ~ "Moderate",
      depressie_percentile == 4 ~ "High",
      depressie_percentile == 5 ~ "Very High",
      TRUE ~ NA_character_
    )
  )

# Optional: View new groupings for first few rows (remove in final report)
amsterdam_gezondheid %>%
  select(wijken_en_buurten, moeite_met_rondkomen, rondkomen_group,
         risico_depressie, depressie_group) %>%
  head(10)

```

## 3.3 Visualize temporal variation

```{r visualise_temporal_variation}

ggplot(amsterdam_gezondheid, aes(x = Perioden)) +
  geom_line(aes(y = avg_risico_depressie, color = "Depressie"), size = 1) +
  geom_line(aes(y = avg_moeite_rondkomen, color = "Rondkomen"), size = 1) +
  facet_wrap(~Leeftijd) +
  labs(title = "Ontwikkeling per leeftijdsgroep", x = "Jaar", y = "Gemiddelde score", color = "Variabele") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6))

```

## 3.4 Visualize spatial variation

```{r visualise_map}

# Load spatial data
geo_data <- st_read("C:/Users/AdhamArtimi/OneDrive/Bureaublad/wijkenbuurten_2022_v3.gpkg", layer = "buurten")

# Filter for Amsterdam only
amsterdam_map <- geo_data |>
  filter(gemeentenaam == "Amsterdam") |>
  select(buurtcode, buurtnaam)

# Clean and filter your health dataset to match buurtcodes only. Here we remove extra spaces in column  'regio_code' and keep only rows that start with "BU"
amsterdam_gezondheid <- amsterdam_gezondheid |>
  mutate(regio_code = toupper(trimws(regio_code))) |>
  filter(str_starts(regio_code, "BU"))

# Merge spatial and health data by buurtcode
amsterdam_merged <- left_join(amsterdam_map, amsterdam_gezondheid, by = c("buurtcode" = "regio_code"))

# Filter to only 2022 data
latest_year <- amsterdam_merged |>
  filter(format(Perioden, "%Y") == "2022")

# Use static mode for PDF
tmap_mode("plot")

# Map 1: Mental health complaints
map1 <- tm_shape(latest_year) +
  tm_polygons("risico_depressie",
              palette = "PuBuGn",
              title = "Psych Complaints (%)") +
  tm_layout(legend.outside = TRUE)

# Map 2: Financial difficulty
map2 <- tm_shape(latest_year) +
  tm_polygons("moeite_met_rondkomen",
              palette = "YlOrRd",
              title = "Financial Stress (%)") +
  tm_layout(legend.outside = TRUE)

# Combine maps
tmap_arrange(map1, map2)


```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}

ggplot(amsterdam_gezondheid, aes(x = Perioden, , y = risico_depressie, fill = Leeftijd)) +
  geom_boxplot(aes(x = Perioden, y = risico_depressie, fill = Leeftijd))

#df2 <- amsterdam_gezondheid |>
#  pivot_longer(cols = c("risico_depressie", "moeite_met_rondkomen"), names_to = "Variabele", values_to = "waarden")

#ggplot(df2, aes(x = as.numeric(Perioden), , y = waarden, fill = Variabele)) +
#  geom_boxplot(aes(x = as.numeric(Perioden), y = waarden, fill = Variabele))

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

ggplot(amsterdam_gezondheid, aes(x = Perioden)) +
  geom_line(aes(y = avg_risico_depressie,  color = "Depressie"), size = 1,) +
  geom_line(aes(y = avg_moeite_rondkomen, color = "Rondkomen"), size = 1) +
  geom_vline(aes(xintercept = as.Date("2020-03-01"), colour = "COVID-19"), linetype = "dashed") +
  labs(title = "Event studie 'COVID-19 2020'", x = "Jaar", y = "Gemiddelde score", color = "Variabele") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6)) +
  scale_color_manual(name = "Variabele",
                     values = c("Depressie" = "Darksalmon", "Rondkomen" = "Darkcyan", "COVID-19" = "Black"))

# https://www.rijksoverheid.nl/onderwerpen/coronavirus-tijdlijn/maart-2020-maatregelen-tegen-verspreiding-coronavirus
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings
Our findings reveal a concerning pattern; areas with higher levels of financial difficulty tend to report elevated rates of psychological complaints. This aligns with the existing literature concerning the relationship between socioeconomic status and mental health. Financial insecurity can therefore be seen as a strong indicator for mental health risks (Marmot, 2005; Pickett& Wilkinson, 2001)

The choropleth maps provide a visual evidence for the spatial disparaties in financial hardship and psychological stress. Several neighbourhoods in Amsterdam, especially those located further from the city center or with historically lower average incomes, show higher levels in both variable consistently.

Sub-group analysis shows that working age aduls (18-65) consitently report the highest levels of both psychological complaints and financial stress. A conclusion could therefore be drawn that economic and mental health pressures are more severe for those in the labour force.

Finally, the event study confirms the strong correlation between financial and psychological strains, following the linear regression line. This supports the idea that economic hardship is closely linked to mental health problems.

These findings align with the public health research and might indicate that addressing financial difficulties locally could be a more effective way to reduce mental health problems across the city.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: https://github.com/Adhamartimi/ProgrammingForEcon-Tutorial-3-group-4

## 5.2 Reference list

Literature: 
Igelström, E., Katikireddi, S. V., Leyland, A., McCartney G., Purba, A. K, Reeves, A., Shimonovich, M., Thomson, H., & Thomson, R. M. (2023). How do income changes impact on mental health and wellbeing for working-age adults? A systematic review and meta-analysis. https://pmc.ncbi.nlm.nih.gov/articles/PMC7614874/#R4 

Marmot, M. (2005): Social determinants of health inequalities. The Lancet. https://www.sciencedirect.com/science/article/pii/S0140673605711466?via%3Dihub 

World Health Organization. (2025). Mental Health. https://www.who.int/health-topics/mental-health#tab=tab_2

Pickett, K. E., & Wilkinson, R. G. (2001). Income inequality and health: A casual review. Social Science & Medicine. https://www.sciencedirect.com/science/article/pii/S0277953614008399?via%3Dihub